The \texttt{mc2d} package.

### 2010-05-18 pkg/mc2d/inst/doc/docmcEnglish.rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/riskassessment@95 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
% sweave commands for vignette
%\VignetteIndexEntry{Tools for Two-Dimensional Monte-Carlo Simulations}
%\VignettePackage{mc2d}
%\VignetteDepends{mvtnorm}

@@

\begin{enumerate}
\item The parameters of the model should be divided into four categories:
the fix parameters, the parameters whose distributions reflect variability only,
hereinafter denoted as \emph{variable parameters}, the parameters
whose distributions reflect uncertainty only, denoted as \emph{uncertain
parameters} and the parameters whose distributions reflect both
uncertainty and variability, denoted as \emph{variable and uncertain parameters}.
For this latter category, a hierarchical
structure, using hyper-parameters, has to be specified: if
a parameter is both uncertain and variable, one should be able to
specify an empirical or parametric distribution representing variability.
This distribution is conditional upon other parameters to which is
associated the uncertainty. As an example, one should be able
to specify a relationship such as
\[ R\left|a,b\right.\sim N\left(a,b\right) \]
where the specified normal distribution represents variability in
$r$ conditional upon parameters $a$ and $b$. Hyperdistributions,
such as
\[ a\sim Unif\left(l_{a},u_{a}\right) \]
and
\[ b\sim Unif\left(l_{b},u_{b}\right) \]
represent the uncertainty in the parameters $a$ and $b$ with uniform distribution;
\item A set of uncertain parameters are randomly sampled from their respective
distributions; 
\item The model is evaluated using a classical (one-dimensional) Monte-Carlo
simulation of size $N_{v}$, treating the uncertain parameters as
fixed. This QRA takes into account the variability in all variable
parameters, and leads to an empirical density function reflecting
the variability of exposure/risk across the population, conditional
upon the uncertain parameters. Various statistics (\emph{e.g.} the
mean, the standard deviation, percentiles) of the resulting empirical
density function are evaluated and stored; 
\item Steps 2) and 3) are performed a large number ($N_{u}$) of times ,
leading to $N_{u}$ sets of statistics. The uncertain parameters are generated
with respect to their uncertain distributions; 
\item As output, the $50^{th}$ percentile (median) of each statistic
is used as a point estimate of this statistic; the $2.5^{th}$
and $97.5^{th}$ percentiles of each statistic are used to
establish a 95\% credible interval (CI95) of this statistic. The median
of the $N_{u}$ estimated values for each of the 101 estimated percentiles
allows us to display a variability cumulative distribution
via a graph. This curve is surrounded by the 2.5th and 97.5th percentiles
obtained from the $N_{u}$ estimates of each of the 101 percentiles.
\end{enumerate}

@@

@@

As a first step, we assume that no uncertainty exists in our model.
All distributions represent variability only. The model may be written
as:
\begin{eqnarray*}
c & = & 10.\\
i & \sim & emp\left(\left\{ 1,1/5,1/50\right\} ,\left\{ 0.027,0.373,0.600\right\} \right)\\
s & \sim & gamma\left(3.93,0.0806\right)\\
n & \sim & Poisson\left(c\times i\times s\right)\\
r & = & 0.001\\
P & = & 1-\left(1-r\right)^{n}
\end{eqnarray*}

where $emp\left(X,P\right)$ is an empirical
distribution wherein each value $X_{i}$ is associated with a probability
$P_{i}$.We will use a classical one
dimensional Monte-Carlo simulation, with 1,000 iterations. Using the
\texttt{mc2d} package, the model may be written as:

@@

The \texttt{print} output provides for each \texttt{node}: 
the \texttt{mode} of the variable (numeric or logical), 
\texttt{nsv}, the size of the variability dimension (here, 1000), 
\texttt{nsu}, the size of the uncertainty dimension (here, 1 since no uncertainty is considered),
\texttt{nva}, the number of variate (here, all the variables are univariate),
some statistics such as the \texttt{min}imum, the \texttt{mean}, 
the \texttt{median} and the \texttt{max}imum,
evaluated after gathering the two dimensions,
the number of missing data \texttt{Nas}, 
the \texttt{type} of the variable (\texttt{0} for fix,
\texttt{V} for variable, \texttt{U} for uncertain or \texttt{VU} for variable
and uncertain (here, all nodes are variable parameters only)
and, finally, the level of output \texttt{outm}, for multivariate parameters.
The \texttt{summary} output proovides additionnal statistics.

@@

We will use a two-dimensional Monte-Carlo simulation, with 1,000 iterations
in the variability dimension and 100 iterations in the uncertainty
dimension. Using the \texttt{mc2d} package, the model may
be written as:

%true
<<sh2, eval=true, echo=true>>=
ndunc(100)
conc <- mcstoc(rnorm,type="U",mean=10,sd=2)
cook <- mcstoc(rempiricalD, type="V",values=c(1,1/5,1/50), prob=c(0.027,0.373,0.600))
serving <- mcstoc(rgamma,type="V",shape=3.93,rate=0.0806)
expo <- conc * cook * serving
dose <- mcstoc(rpois,type="VU",lambda=expo)
r <- mcstoc(runif,type="U",min=0.0005,max=0.0015)
risk <- 1-(1-r)^dose
EC2 <- mc(conc,cook,serving,expo,dose,r,risk)
print(EC2,digits=2)
summary(EC2)
@

@@

The summary provides estimates of the variability distributions (in
rows) but with a measure of their uncertainty, linked to the uncertainty
around \texttt{conc} and \texttt{r}. The estimate of the mean risk
is now uncertain. The median of the 100 simulations leads to a best
estimateof 0.0445, with a 95\% credible
interval of {[}0.191, 0.0730{]}.

@@

An \texttt{mcnode} is an array of dimension \texttt{nsv} * \texttt{nsu} * \texttt{nvariates}
where $nsv$ is the dimension of variability, $nsu$ is the dimension
of uncertainty and $nvariates$ is the number of variates of the \texttt{mcnode}%
\footnote{In this section, we will only consider univariate \texttt{mcnode}s,
that is \texttt{mcnode}s with $nvariates=1$.%
}. Four types of \texttt{mcnode} exist: 
\begin{itemize}
\item \texttt{V} \texttt{mcnode}, for \emph{Variability},
is an array of dimension $\left(nsv\times1\times nvariates\right)$.
The distribution represents variability in the parameter; 
\item \texttt{U} \texttt{mcnode}, for \emph{Uncertainty},
is an array of dimension $\left(1\times nsu\times nvariates\right)$.
The distribution represents uncertainty in the parameter. 
\item \texttt{VU} \texttt{mcnode}, for \emph{Variability
and Uncertainty}, is an array of dimension $\left(nsv\times nsu\times nvariates\right)$.
The distribution represents both variability (in the first dimension)
and uncertainty (in the second dimension) in the parameter. 
\item Additionally, a \texttt{0} \texttt{mcnode}
is also defined . \texttt{0} stands
for Neither Variability or Uncertainty.
Such nodes are arrays of dimension $\left(1\times1\times nvariates\right)$.
No uncertainty or variability is considered for these nodes. A \texttt{0}
\texttt{mcnode} is not necessary in the univariate context (use a
scalar instead) but is useful in constructing multivariate nodes (See
section \ref{sec:Multivar}). 
\end{itemize}
%
\begin{figure}
\noindent \begin{centering}
\includegraphics[angle=270, width=.7\textwidth]{Illustration}
\par\end{centering}

@@

Structural links between sets of parameters may be very important
in QRA. In \texttt{mc2d}, a Spearman rank correlation structure for
2 or more nodes may be specified with the \texttt{cornode} function.
This function uses the method of Iman \& Conover to generate correlated
samples \cite{IMAN-CONOVER-1982}. Assume that a study suggests that
people who consume rare ground beef also consume larger serving sizes.
We could specify this relation using:

@@

\footnote{Note that the resulting correlation (around 0.4) is obviously an approximation
to the desired value of 0.5, because a discrete distribution (\texttt{cook}:
3 categories) is correlated with a continuous distribution (\texttt{serving}).}

@@

\noindent \begin{flushleft}
\texttt{plot (x, prec = 0.01, stat = c(median,
mean), lim = c(0.025,0.25, 0.75, 0.975), na.rm =
TRUE, griddim = NULL, xlab = NULL, ylab = Fn(x),
main = , draw = TRUE, ...)}
\par\end{flushleft}

For our example, see Figure \ref{fig:Function-plot}, a default graph.

@@

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
% sweave commands for vignette
%\VignetteIndexEntry{Tools for Two-Dimensional Monte-Carlo Simulations}
%\VignettePackage{mc2d}
%\VignetteDepends{mvtnorm,fitdistrplus}

@@

\begin{enumerate}
\item The parameters of the model should be divided into four categories:
the fix parameters, the parameters whose distributions reflect variability only,
hereinafter denoted as \emph{variable parameters}, the parameters
whose distributions reflect uncertainty only, denoted as \emph{uncertain
parameters} and the parameters whose distributions reflect both
uncertainty and variability, denoted as \emph{variable and uncertain parameters}.
For this latter category, a hierarchical
structure, using hyper-parameters, has to be specified: if
a parameter is both uncertain and variable, one should be able to
specify an empirical or parametric distribution representing variability.
This distribution is conditional upon other parameters to which is
associated the uncertainty. As an example, one should be able
to specify a relationship such as
\[ r\left|a,b\right.\sim N\left(a,b\right) \]
where the specified normal distribution represents variability in
$r$ conditional upon parameters $a$ and $b$. Hyperdistributions,
such as
\[ a\sim Unif\left(l_{a},u_{a}\right) \]
and
\[ b\sim Unif\left(l_{b},u_{b}\right) \]
represent the uncertainty in the parameters $a$ and $b$ with uniform distribution;
\item A set of uncertain parameters are randomly sampled from their respective
distributions; 
\item The model is evaluated using a classical (one-dimensional) Monte-Carlo
simulation of size $N_{v}$, treating the uncertain parameters as
fixed. This QRA takes into account the variability in all variable
parameters, and leads to an empirical density function reflecting
the variability of exposure/risk across the population, conditional
upon the uncertain parameters. Various statistics (\emph{e.g.} the
mean, the standard deviation, percentiles) of the resulting empirical
density function are evaluated and stored; 
\item Steps 2) and 3) are performed a large number ($N_{u}$) of times ,
leading to $N_{u}$ sets of statistics. The uncertain parameters are generated
with respect to their uncertain distributions; 
\item As output, the $50^{th}$ percentile (median) of each statistic
is used as a point estimate of this statistic; the $2.5^{th}$
and $97.5^{th}$ percentiles of each statistic are used to
establish a 95\% credible interval (CI95) of this statistic. The median
of the $N_{u}$ estimated values for each of the 101 estimated percentiles
allows us to display a variability cumulative distribution
via a graph. This curve is surrounded by the 2.5th and 97.5th percentiles
obtained from the $N_{u}$ estimates of each of the 101 percentiles.
\end{enumerate}

@@

\emph{Warning}: the data are fictitous and the model is over-simplified 
to better illustrate the use of the package: 
the results will not and should not be interpreted.

@@

As a first step, we assume that no uncertainty exists in our model.
All distributions represent variability only. The model may be written
as:
\begin{eqnarray*}
c & = & 10.\\
i & \sim & emp\left(\left\{ 1,1/5,1/50\right\} ,\left\{ 0.027,0.373,0.600\right\} \right)\\
s & \sim & gamma\left(3.93,0.0806\right)\\
n & \sim & Poisson\left(c\times i\times s\right)\\
r & = & 0.001\\
P & = & 1-\left(1-r\right)^{n}
\end{eqnarray*}
where $emp\left(X,P\right)$ is an empirical
distribution wherein each value $X_{i}$ is associated with a probability
$P_{i}$.We will use a classical one
dimensional Monte-Carlo simulation, with 1,001 iterations. Using the
\texttt{mc2d} package, the model may be written as:

@@

The \texttt{print} output provides for each \texttt{node}: 
the \texttt{mode} of the variable (numeric or logical), 
\texttt{nsv}, the size of the variability dimension (here, 1001), 
\texttt{nsu}, the size of the uncertainty dimension (here, 1 since no uncertainty is considered),
\texttt{nva}, the number of variate (here, all the variables are univariate),
some statistics such as the \texttt{min}imum, the \texttt{mean}, 
the \texttt{median} and the \texttt{max}imum,
evaluated after gathering the two dimensions,
the number of missing data \texttt{Nas}, 
the \texttt{type} of the variable (\texttt{0} for fix,
\texttt{V} for variable, \texttt{U} for uncertain or \texttt{VU} for variable
and uncertain (here, all nodes are variable parameters only)
and, finally, the level of output \texttt{outm} (used for multivariate parameters).
The \texttt{summary} output proovides additionnal statistics.

@@

We will use a two-dimensional Monte-Carlo simulation, with 1,001 iterations
in the variability dimension and 101 iterations in the uncertainty
dimension. Using the \texttt{mc2d} package, the model may
be written as:

%true
<<sh2, eval=true, echo=true>>=
ndunc(101)
conc <- mcstoc(rnorm,type="U",mean=10,sd=2)
cook <- mcstoc(rempiricalD, type="V",values=c(1,1/5,1/50), prob=c(0.027,0.373,0.600))
serving <- mcstoc(rgamma,type="V",shape=3.93,rate=0.0806)
expo <- conc * cook * serving
dose <- mcstoc(rpois,type="VU",lambda=expo)
r <- mcstoc(runif,type="U",min=0.0005,max=0.0015)
risk <- 1-(1-r)^dose
EC2 <- mc(conc,cook,serving,expo,dose,r,risk)
print(EC2,digits=2)
summary(EC2)
@

@@

The summary provides estimates of the variability distributions (in
rows) but with a measure of their uncertainty, linked to the uncertainty
around \texttt{conc} and \texttt{r}. The estimate of the mean risk
is now uncertain. The median of the 101 simulations leads to a best
estimate of 0.0445, with a 95\% credible
interval of {[}0.191, 0.0730{]}.

@@

An \texttt{mcnode} is an array of dimension \texttt{nsv x nsu x nvariates}
where $nsv$ is the dimension of variability, $nsu$ is the dimension
of uncertainty and $nvariates$ is the number of variates of the \texttt{mcnode}%
\footnote{In this section, we will only consider univariate \texttt{mcnode}s,
that is \texttt{mcnode}s with $nvariates=1$.%
}. Four types of \texttt{mcnode} exist: 
\begin{itemize}
\item \texttt{V} \texttt{mcnode}, for \emph{Variability},
is an array of dimension $\left(nsv\times1\times nvariates\right)$.
The distribution represents variability in the parameter; 
\item \texttt{U} \texttt{mcnode}, for \emph{Uncertainty},
is an array of dimension $\left(1\times nsu\times nvariates\right)$.
The distribution represents uncertainty in the parameter. 
\item \texttt{VU} \texttt{mcnode}, for \emph{Variability
and Uncertainty}, is an array of dimension $\left(nsv\times nsu\times nvariates\right)$.
The distribution represents both variability (in the first dimension)
and uncertainty (in the second dimension) in the parameter. 
\item Additionally, a \texttt{0} \texttt{mcnode}
is also defined . \texttt{0} stands
for Neither Variability or Uncertainty.
Such nodes are arrays of dimension $\left(1\times1\times nvariates\right)$.
No uncertainty or variability is considered for these nodes. A \texttt{0}
\texttt{mcnode} is not necessary in the univariate context (use a
scalar instead) but is useful in constructing multivariate nodes (See
section \ref{sec:Multivar}). 
\end{itemize}
%
\begin{figure}
\noindent \begin{centering}
\includegraphics[angle=270, width=.7\textwidth]{Illustration}
\par\end{centering}

@@

Structural links between sets of parameters may be very important
in QRA. In \texttt{mc2d}, a Spearman rank correlation structure for
2 or more nodes may be specified with the \texttt{cornode} function.
This function uses the method of Iman \& Conover to generate correlated
samples \cite{IMAN-CONOVER-1982}. Assume that a study suggests that
people who consume rare ground beef also consume larger serving sizes.
We could specify this relation using\footnote{Note that the resulting correlation (around 0.4) is obviously an approximation
to the desired value of 0.5, because a discrete distribution (\texttt{cook}:
3 categories) is correlated with a continuous distribution (\texttt{serving}).}:

@@

@@

\noindent \begin{flushleft}
\texttt{plot(x, prec = 0.001, stat = c("median", "mean"), lim = c(0.025, 
    0.25, 0.75, 0.975), na.rm = TRUE, griddim = NULL, xlab = NULL, 
    ylab = "Fn(x)", main = "", draw = TRUE, paint = TRUE, ...)}
\par\end{flushleft}

For our example, see Figure \ref{fig:Function-plot}, a default graph.
The 0.25 and 0.75 quantiles (default values of \texttt{lim}) in the uncertainty dimension
of the quantiles (variability dimension) are used as the enveloppe.

@@

\subsubsection{The \texttt{mcratio} function}

The \texttt{mcratio} function provides measures of variability, uncertainty, 
and both combined propose by \cite{OZKAYNAK-2009} for an \texttt{mc} or an \texttt{mcnode} object.
Given: 
\begin{description}
\item[A] the median of uncertainty for the median of variability; 
\item[B] the median of uncertainty for the 97.5th percentile of variability; 
\item[C] the 97.5th percentile of uncertainty for the median percentile of variability; 
\item[D] the 97.5th percentile of uncertainty for the 97.5th percentile of variability. 
\end{description}
The following ratio are estimated: 
\begin{itemize}
\item Variability Ratio: B / A 
\item Uncertainty Ratio: C / A 
\item Overall Uncertainty Ratio: D / A
\end{itemize}

%true
<<sh19d, echo=true, eval=true, fig=false>>=
mcratio(risk)
@

@@

\end{document}
<- mcstoc(rbeta, type="U", shape1=2.65, shape2=3.64)
w <- mcstoc(rbeta, type="V", shape1=2.6, shape2=3.4) 
@

<<Crypto11, echo=true, figure=false>>=
Oo <- 2
l <- (Oo + mcstoc(rnbinom, type="U", size=Oo+1, prob=Rr))/100
@

<<Crypto12, echo=true, figure=false>>=
Or <- l * v * w
P <- 10000 * (1-exp(-r*Or))
summary(P)
@

This result can be compared (roughly since there is some differences
in the modelsvariability) to the results shown in Table 2 in \cite{POUILLOT-2004}.

We think and hope that {}``\texttt{mc2d}'' could help risk assessors
to constuct and analyse their models, and that it may help in developing
\textquotedbl{}two-dimensional\textquotedbl{} simulations. Nevertheless,
\textquotedbl{}\texttt{mc2d}\textquotedbl{} is currently under development:

\noindent \begin{center}
\emph{CHECK YOUR MODEL CAREFULLY AND EXAMINE RESULTS TO DETECT BUGS}
\par\end{center}

and, if you would like to improve it, join us at

\noindent \begin{center}
\emph{\url{http://riskassessment.r-forge.r-project.org/}}
\par\end{center}

Please refer any comments or bugs to \url{rpouillot@yahoo.fr}.

\bibliographystyle{plain}
\bibliography{Bibtex}
